Asymmetric relationships and their effects on coexistence

### 2023-07-28 paper/Paper.Rnw
took a stab at a Discussion paragraph

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant through time. 
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the whole paper is summarized in Table \ref{tab:notation}.
Typesetting procedures used by publishers of ecological journals render mathematics sub-optimally,
so we have appended a version of this main text rendered using latex to the end of the 
Supporting Information (SI); we recommend that readers interested in the mathematical content
of the study read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, with a fuller understanding
necessarily coming only later in the paper.
It is known that storage effects can occur if some buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
%
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
%
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal.
Depending on the nature of population growth, storage potential may differ
across the spectrum from 1 to 2.
%
ATA effects are essentially the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are strongly associated in the parts of their distribution where
storage potential is high, ATA effects will be low, and vice versa.

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
%We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
%spatial variation can be considered in future work.
G2) We explore, using the simple lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and we determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs relative to other mechanisms of coexistence.
The example is a diatom system in a chemostat. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
And we return in the Discussion to the task
of building intuition for how ATAs work. 
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

>>>>

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant through time. 
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the whole paper is summarized in Table \ref{tab:notation}.
Typesetting procedures used by publishers of ecological journals render mathematics sub-optimally,
so we have appended a latex version of this main text to the end of the 
Supporting Information (SI). Readers interested in the mathematical content
of the study should read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
providing additional intuition in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high, and vice versa.

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
%We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
%spatial variation can be considered in future work.
G2) We explore, using the simple lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and we determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs relative to other mechanisms of coexistence.
The example is a diatom system in a chemostat. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

%Para continuing to try to build intuition about how ATAs work
We return to the task, begun in the Introduction, of building intuition for how
ATA effects work. 
Just as for storage effects [\cite{Chesson_1994}, Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary for ATA
effects (SI section X1). 
%DAN: fix reference
If temporal niche differentiation occurs
but is partial, as will be typical, $E_i$ and $C_{i \bs i}$ for a 
species $i$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
Storage events occur for species $i$ when the positive effects of $E_i$ overwhelm the negative 
effects of $C_{i \bs i}$, producing large values of $r_i(E_i,C_{i \bs i})$
which are not completely offset by declines at other times. 
For instance, for the lottery model,
$r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$ can get large whenever
$C_{i \bs i}$ is a small fraction of $E_i$, but can never get smaller
than $\log(1-\delta)$. 
The potential for storage events can differ when $E_i$ and $C_{i \bs i}$ are both
above their median values compared to when they are both below median. For instance,
the ratio $E_i/C_{i \bs i}$, and therefore the growth rate 
$\log(1-\delta+E_i/C_{i \bs i})$, can both get very big, corresponding to
rapid growth, if $C_{i \bs i}$ is close to $0$, even if $E_i$ is fairly small.
In contrast, if $E_i$ and $C_{i \bs i}$ are both in the right portions of their distributions, 
so that $C_{i \bs i}$ is bounded away from $0$, there may be much sharper limits on the 
ratio $E_i/C_{i \bs i}$ and the corresponding growth rate.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with varying potential for storage across the $E,C$ distribution. 
ATA effects help quantify different degrees of temporal niche differentiation 
occurring under different sets of environmental conditions, possibly via
different mechanisms.
Details and examples of this intuition 
are in SI section X2, 
%DAN: Fix reference
where comparisons are also made between our beta and log-normal fecundities 
lottery models.
